---
title: "Project"
author: "Acacia, Kate, Julia"
date: "3/23/2021"
output: 
      html_document: default
      pdf_document: default
---

```{r setup, include=FALSE}
set.seed(1)
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)            
```

## Scenario
Genes A and B have both been shown to be related to lifetime chance of cancer. We want to know if an organism with gene A has the same chance of cancer as an organism with gene B, and how that differs when an organism has both genes A and B, or neither gene. We need to include the covariate of the organism’s mass, because we know this also affects the lifetime chance of cancer. 

We sample N organisms of each of the four genotypes, and we record their body mass and whether or not cancer is present in that organism. 

### Hypothesis

_Null Hypothesis_: The probability of cancer is equal across all genotypes.

_Alternative Hypothesis_: The probability of cancer is different between genotypes.

### Variables

_Genotypes_: WT (neither A nor B), A, B, and AB. A and B each have an incidence rate for cancer.

_Mass_: a continuous variable between 1 and 10 grams

_Genotype A x Genotype B_: Interaction effects

_Response variable_: Presence or absence of cancer in the organism (0/1) 

_Predictor variables_: Genotype A, Genotype B, interaction between Genotypes A & B, and organism mass 


## Modeling and Justification


**Binomial Distribution**

$y \sim Bin(p,N)$

- $p$ is probability of cancer
- $N$ is total number of individuals

_Justification_: We measure cancer as either present/ absent in each organism. The data simulation evolves N multicellular individuals. We have fixed N number of trials that we can repeat _ad infinitum_. This is a frequentists' approach. Hence, we chose Binomial distribution.

**Deterministic Function (model) & Joint Probability (likelihood)**

$counts = \alpha + \beta_1 * genotype_A + \beta_2 * genotype_B + \delta * genotype_A * genotype_B + \gamma * mass + \epsilon$

- $counts$ = expected number of organisms with cancer present
- $\alpha$ = (intercept) -- expected incidence rate of cancer in wildtype reference
- $\beta_1$ = difference in mean rate of cancer for genotype_A compared to wildtype (how much gene A influences incidence of cancer)
- $\beta_2$ = difference in mean rate of cancer for genotype_B compared to wildtype (how much gene B influences incidence of cancer)
- $\delta$ = slope term indicating the interaction relationship between genes A and B (how much genes A and B combined influence incidence of cancer)
- $\gamma$ = constant for mass (how much organism mass influences incidence of cancer)
- $\epsilon$ = randomness pulled from binomial distribution
  + $\epsilon \sim Bin(p,N)$

---

## Simulate the Data

```{r}

#### Number of Datapoints ####
# Assume a balanced design
# 2 kinds of genes - A and B
# Arbitrary number of replicates performed
n.reps <- 80 #50   #changed to 80 so that 4 divides evenly into it

#Total number of data points
n <- n.reps

# For a binomial distribution, we need to know the number of organisms in each replicate
# In our work we usually use the same number of organsims in each replicate
n.orgs <- rep(100,n)


#### Data Simulation ####

# Specify a categorical variable which indicates genotype
genA <- factor(rep(c(0,1), each=n/2)) # produces 0000...1111....
genB <- factor(rep(rep(c(0,1), each=n/4),2)) # produces 00...11...00...11...

# We also need a vector to indicate mass for each organism
# Assume this is a continuous variable between 1 and 10 grams
mass <- round(runif(n, min=1, max=10), digits=2)

# Chose the values for the parameters (logit transformed)
# Labeled to make it easier to think about
# Each element in beta.vec.names indicates the effect of that variable or interaction of variables
beta.vec.names <- c("WT", "genA", "genB", "genA:genB", "mass")
beta.vec <- c(-0.4, 0.3, 0.1, 0.6, 0.2) # we can freely edit these
names(beta.vec) <- beta.vec.names


#### Model Matrix Creation ####
# Build the design matrix of the interactive combination of genotype and mass
Xmat = model.matrix(~genA*genB + mass)

#### Create Stochastic Data ####
#Generate the linear predictor (mutliple Xmat by the beta.vec)
lin.pred =  Xmat%*%beta.vec

#Transform the data with an inverse logit to get the expected proportion of cancerous samples
exp.p <- exp(lin.pred)/(1+exp(lin.pred))

# Add binomial noise
cancer.counts <- rbinom(n=n, size=n.orgs, prob=exp.p)

#### Combine Data ####
# Combine type data, mass data, and cancer counts
df <- data.frame(genA, genB, mass, cancer.counts)

#### Export ####
# Export the data to a csv file
filename <- "cancer_data.csv"
write.csv(df, file=filename, row.names = FALSE)
```

---
## Parameter estimation method

---
## Statistical Test

---

## Plots



```{r, echo=FALSE}
```

---

## Analysis
```{r}
```
